NOPT048 Notes

### 2019-09-30 misc/study-notes/groups.tex
Fixed typo in groups and fixed teletype
Possibly fixed issues regarding mathmac and fontmac clashing

The term {\it order} serves a somewhat dual purpose in group theory. The order of a group $G$ is the number of elements it contains and and this value, also denoted $|G|$, need not be finite. On the other hand, we define the order of a group {\it element} $g$ to be the smallest $k>0$ such that $g^k=e$. If no such $k$ exists, then $g$ is said to have infinite order.

@@

\thmparens L {Lagrange} Let $H$ be a subgroup of a group $G$. Then the order of $H$ must divide the order of $G$.\slug

@@

@@

Let $f:G\rightarrow G'$. The {\it kernel} of $f$, denoted $\ker(f)$, is a the set of all elements in the $G$ that map to the identity $e'$ of the target group $G'$. The {\it image} of $f$, denoted $\Im(f)$, is the set of all elements in $G'$ that equal $f(g)$ for some $g\in G$. Then the homomorphism $f$ is an isomorphism if and only if $\ker(f) = \{e\}$ and $\Im(f) = G'$. It is an easy exercise to verify that $\ker(f)$ and $\Im(f)$ are subgroups of $G$ and $G'$ respectively. In fact, the kernel of a homomorphism is a normal subgroup.

>>>>

The term {\it order} serves a somewhat dual purpose in group theory. The order of a group $G$ is the number of elements it contains and this value, also denoted $|G|$, need not be finite. On the other hand, we define the order of a group {\it element} $g$ to be the smallest $k>0$ such that $g^k=e$. If no such $k$ exists, then $g$ is said to have infinite order.

@@

\parenproclaim Theorem L (Lagrange). Let $H$ be a subgroup of a group $G$. Then the order of $H$ must divide the order of $G$.\slug

@@

Now we introduce the concept of conjugacy, which is closely intertwined with normal subgroups. For elements $a$ and $b$ of a group $G$, we say $a$ is {\it conjugate} to $b$ if there exists an element $g\in G$ such that $gag^{-1} = b$. It is easy to see that this relation is reflexive (with $g=e$) and symmetric (replace any $g$ with its inverse), so it makes sense to say that $a$ and $b$ are conjugate elements of a group. We can also prove transitivity: If there exists $g\in G$ such that $gag^{-1} = b$ and $h\in G$ such that $hbh^{-1} = c$, then $hgag^{-1}h^{-1}= (hg)a(hg)^{-1} = c$. Thus conjugacy is an equivalence relation that partitions a group into {\it conjugacy classes}.

The following lemma describes the link between conjugacy and normal subgroups.

\proclaim Lemma C. Normal subgroups are exactly the subgroups that are stable under conjugation.

\proof Let $H$ be a normal subgroup of a group $G$. Let $g\in G$ and $h\in H$ be arbitrarily chosen. Since $gH=Hg$, there is an element $h'\in H$ such that $gh = h'g$. Then we see that $ghg^{-1} = h'$, an element in $H$. So we conclude that normal subgroups are stable under conjugation.

On the other hand, suppose $H$ is a subgroup such that, for any $g\in G$ and $h\in H$, $ghg^{-1}\in H$. Then $gHg^{-1}\subset H$ and we can conclude that $gH\subset Hg$. But $g^{-1}$ is also an element of $G$, so we can apply our hypothesis to get that $g^{-1}Hg\subset H$ and $Hg \subset gH$. So $gH=Hg$ for all $g\in G$ and $H$ is normal.\slug

@@

Let $f:G\rightarrow G'$. The {\it kernel} of $f$, denoted $\ker(f)$, is a the set of all elements in the $G$ that map to the identity $e'$ of the target group $G'$. The {\it image} of $f$, denoted $\Im(f)$, is the set of all elements in $G'$ that equal $f(g)$ for some $g\in G$. Then the homomorphism $f$ is an isomorphism if and only if $\ker(f) = \{e\}$ and $\Im(f) = G'$. It is an easy exercise to verify that $\ker(f)$ and $\Im(f)$ are subgroups of $G$ and $G'$ respectively. In fact, the kernel of a homomorphism is a normal subgroup, as we shall now prove.

\proclaim Theorem K. Let $f:G\rightarrow G'$ be a group homomorphism. Then $K = \ker(f)$ is a normal subgroup of $G$.

\proof By Lemma 2.4.C, we need only show that $K$ is stable under conjugation. Let $k\in K$ be given and let us conjugate it by an element $a\in G$ to get $aka^{-1}$. Applying the homomorphism $f$ to $aka^{-1}$, we find that
$$f(aka^{-1}) = f(a)f(k)f(a^{-1}) = f(a)e'f(a)^{-1},$$
where $e'$ is the identity of $G'$. But $f(a)e'f(a)^{-1} = e'$, so $aka^{-1}\in \ker(f) = K$.\slug

### 2019-09-30 misc/study-notes/ndmi011notes.tex
Fixed typo in graph theory and added algebra
Added flairs
Possibly fixed issues regarding mathmac and fontmac clashing

\thmparens{G}{Generalised binomial theorem} For any $r\in \R$,
$$\sum_{i\geq 0} {r\choose i} x^i = (1 + x)^r.$$

@@

\thmparens{H}{Hall} A bipartite graph $G=(A\cup B, E)$ (with $A$ and $B$ disjoint) has a matching $M\subseteq E$ of size $|M| = |A|$ if and only if $|N(S)|\geq |S|$ for all $S\subseteq A$.

@@

\thmparens{K}{K\H onig's theorem} Let $G$ be a bipartite graph with maximum matching $M$ and minimum vertex cover $C$. Then $|M| = |C|$.\slug

@@

\thmparens{M}{Menger}
Let $G$ be a graph. The following both hold:
\medskip
\item {1.} $G$ is $k$-connected if and only if there are $k$ independent paths between any two vertices $s$ and $t$.
\smallskip
\item {2.} $G$ is $k$-edge-connected if and only if there are $k$ edge-disjoint paths between any two vertices $s$ and $t$.
\medskip

@@

\thmparens{C}{Cayley's formula} The number of spanning trees of a complete graph with $n$ vertices is
$$ T(K_n) = n^{n-2}.$$

@@

\thmparens{S}{Sperner} Any antichain of a poset of $X = \{1, \ldots, n\}$ has size at most $n\choose \lfloor n/2 \rfloor$

@@

\proclaim Theorem R. Let $G=(V,E)$ be a graph. If $|V|\geq {k-l-2\choose k-1} = {k+l-2\choose l-1}$, then $\alpha(G)\geq l$ or $\omega(G)\geq k$.

\proof By induction on $k+l$. In the base case $k = 1$ or $l=1$, all the theorem says is that there is an independent set or clique of size 4. For the inductive step, let $k,l\geq 2$ and assume that the claim holds for $k,l-1$ or $k-1,l$. By Pascal's formula,
$${k+l-2\choose k-1} = {k+l-3\choose k-1} + {k+l-3\choose k-2}.$$
Call the left-hand side $n$ and from the right-hand side, call the first summand $n_1$ and the second summand $n_2$. Now pick any vertex $u\in V$. Let $B = N(u)$ and $A = V\setminus(B\cup \{u\})$. First note that it is impossible that both $|A| < n_1$ and $|B|<n_2$ since that would imply that
$$n = 1 + |A| + |B| \leq 1 + (n_1-1) + (n_2-1) = n-1,$$
a contradiction. So either $|A|\geq n_1$ or $|B|\geq n_2$.

@@

\thmparens{E}{Erd\H os-Szekeres} For any finite sequence $S$ of at least $(r-1)(s-1) + 1$ numbers, there is an increasing subsequence of length $r$ or a decreasing subsequence of length $s$.

@@

\thmparens{H}{Hamming bound} If an $(n,k,d)_q$-code exists, then
$$q^k\cdot\hbox{\rm vol}\left(\left\lfloor {d-1\over 2} \right\rfloor, n\right) \leq q^n.$$

@@

\thmparens{S}{Singleton bound} If $C$ is an $(n,k,d)_q$-code, then $k\leq n-d+1$.

>>>>

\parenproclaim Theorem G (Generalised binomial theorem). For any $r\in \R$,
$$\sum_{i\geq 0} {r\choose i} x^i = (1 + x)^r.$$

@@

\parenproclaim Theorem H (Hall). A bipartite graph $G=(A\cup B, E)$ (with $A$ and $B$ disjoint) has a matching $M\subseteq E$ of size $|M| = |A|$ if and only if $|N(S)|\geq |S|$ for all $S\subseteq A$.

@@

\parenproclaim Theorem K (K\H onig's theorem). Let $G$ be a bipartite graph with maximum matching $M$ and minimum vertex cover $C$. Then $|M| = |C|$.\slug

@@

\parenproclaim Theorem M (Menger). Let $G$ be a graph. The following both hold:
\medskip
\item {1.} $G$ is $k$-connected if and only if there are $k$ independent paths between any two vertices $s$ and $t$.
\smallskip
\item {2.} $G$ is $k$-edge-connected if and only if there are $k$ edge-disjoint paths between any two vertices $s$ and $t$.
\medskip

@@

\parenproclaim Theorem C (Cayley's formula). The number of spanning trees of a complete graph with $n$ vertices is
$$ T(K_n) = n^{n-2}.$$

@@

\parenproclaim Theorem S (Sperner). Any antichain of a poset of $X = \{1, \ldots, n\}$ has size at most $n\choose \lfloor n/2 \rfloor$

@@

\proclaim Theorem R. Let $G=(V,E)$ be a graph. If $|V|\geq {k+l-2\choose k-1} = {k+l-2\choose l-1}$, then $\alpha(G)\geq l$ or $\omega(G)\geq k$.

\proof By induction on $k+l$. In the base case $k = 1$ or $l=1$, all the theorem says is that there is an independent set or clique of size 1. For the inductive step, let $k,l\geq 2$ and assume that the claim holds for $k,l-1$ or $k-1,l$. By Pascal's formula,
$${k+l-2\choose k-1} = {k+l-3\choose k-1} + {k+l-3\choose k-2}.$$
Call the left-hand side $n$ and from the right-hand side, call the first summand $n_1$ and the second summand $n_2$. Now pick any vertex $u\in V$. Let $B = N(u)$ and $A = V\setminus(B\cup \{u\})$. First note that it is impossible that both $|A| < n_1$ and $|B|<n_2$ since that would imply that
$$n = 1 + |A| + |B| \leq 1 + (n_1-1) + (n_2-1) = n-1,$$
a contradiction. So either $|A|\geq n_1$ or $|B|\geq n_2$.

@@

\parenproclaim Theorem E (Erd\H os-Szekeres). For any finite sequence $S$ of at least $(r-1)(s-1) + 1$ numbers, there is an increasing subsequence of length $r$ or a decreasing subsequence of length $s$.

@@

\parenproclaim Theorem H (Hamming bound). If an $(n,k,d)_q$-code exists, then
$$q^k\cdot\hbox{\rm vol}\left(\left\lfloor {d-1\over 2} \right\rfloor, n\right) \leq q^n.$$

@@

\parenproclaim Theorem S (Singleton bound). If $C$ is an $(n,k,d)_q$-code, then $k\leq n-d+1$.

### 2019-09-30 misc/study-notes/nopt048notes.tex
Possibly fixed issues regarding mathmac and fontmac clashing

\thmparens{M}{Minkowski-Weyl} A set of points $P$ is a polyhedron if and only if there exist finite sets $V$ and $Y$ such that $P = \conv V + \cone Y$ where the sum $A + B$ denotes the set $\{a + b : a\in A, b\in B\}$.

@@

\thmparens{W}{Weak Duality Theorem} Let $P$ and $D$ be primal and dual linear programs of the above form and let $x$ and $y$ be respective feasible solutions. Then $c^Tx \leq b^Ty$.

\thmparens{S}{Strong Duality Theorem} Let $P$ and $D$ be primal and dual linear programs. Then exactly one of the following holds:
\medskip
\item {i)} P infeasible, D infeasible
\smallskip
\item {ii)} P infeasible, D unbounded
\smallskip
\item {iii)} P unbounded, D infeasible
\smallskip
\item {iv)} P and D both feasible and bounded and furthermore, if $x^*$ and $y^*$ are primal and dual optimal solutions, then $c^Tx^* = b^Ty^*$.
\medskip

\lemparens{F}{Farkas lemma}
For an arbitrary matrix $A$ and vector $b$, exactly one of the following holds:
\medskip
\item {i)} There exists a vector $x$ such that $Ax = b$ and $x \geq 0$.
\smallskip
\item {ii)} There exists a vector $y$ such that $A^Ty \geq 0$ and $b^Ty< 0$.
\medskip

\lemparens{G}{Farkas lemma 2.0}
For an arbitrary matrix $A$ and vector $b$, exactly one of the following holds:
\medskip
\item {i)} There exists a vector $x$ such that $Ax \leq b$ and $x \geq 0$.
\smallskip
\item {ii)} There exists a vector $y$ such that $A^Ty \geq 0$, $y\geq0$ and $b^Ty< 0$.
\medskip

@@

Given a graph $G=(V,E)$, a {\it matching} is a set $M\subset E$ such that for all $v\in V$ there is at most one edge $e\in M$ with $v\in e$. The linear program that finds the maximal matching in a bipartite graph is
$$\matrix{
    & \maximise   & \quad & \displaystyle\sum_{e\in E} x_e  \hfill    & \hcr
    & \subject &       & \displaystyle\sum_{v\in e}x_e \leq 1  \hfill &\hbox{for all}\ v\in V,\hcr
    &                   &       & x_e \geq 0      \hfill          &\hbox{for all}\ e\in E.\hfill
}$$
The dual of this linear program is the following:
$$\matrix{
    & \minimise   & \quad & \displaystyle\sum_{v\in V} y_v  \hfill & \hcr
    & \subject &       & y_u + y_v \geq 1   \hfill &\hbox{for all}\ \{u,v\}\in E,\hcr
    &                   &       & y_v \geq 0        \hfill  &\hbox{for all}\ v\in V.\hfill
}$$
    Given a graph $G=(V,E)$, a {\it vertex cover} is a set $S \subset V$ such that for all $\{u, v\}\in E$, $\{u,v\}\cap S\geq 1$. The dual linear program above finds the minimum vertex cover in a bipartite graph. The duality of these two programs results in K\H{o}nig's theorem.
\thmparens{K}{K\H{o}nig's theorem}
    Suppose $G=(V,E)$ is a bipartite graph. Let $M\subset E$ denote the maximum matching and let $S\subset V$ denote the minimum vertex cover. Then $|M|=|S|$.\slug

@@

\thmparens{R}{Rank characterisation of matroids} Let $E$ be a finite set and suppose that $r : {\cal P}(E)\rightarrow \R$ satisfies the three peoperties of the rank function. Then $M=(E,I)$, with
$$I = \{Y \subset E : |Y| = r(Y)\},$$
is a matroid with $r$ as its rank function.

@@

\thmparens{E}{Greedy equivalence} Let $M=(E,I)$ be an independence system. Then if the weighted greedy algorithm works for all (non-negative) weights, then $M$ is a matroid.\slug

>>>>

\parenproclaim Theorem M (Minkowski-Weyl). A set of points $P$ is a polyhedron if and only if there exist finite sets $V$ and $Y$ such that $P = \conv V + \cone Y$ where the sum $A + B$ denotes the set $\{a + b : a\in A, b\in B\}$.

@@

\parenproclaim Theorem W (Weak Duality Theorem). Let $P$ and $D$ be primal and dual linear programs of the above form and let $x$ and $y$ be respective feasible solutions. Then $c^Tx \leq b^Ty$.

\parenproclaim Theorem S (Strong Duality Theorem). Let $P$ and $D$ be primal and dual linear programs. Then exactly one of the following holds:
\medskip
\item {i)} P infeasible, D infeasible
\smallskip
\item {ii)} P infeasible, D unbounded
\smallskip
\item {iii)} P unbounded, D infeasible
\smallskip
\item {iv)} P and D both feasible and bounded and furthermore, if $x^*$ and $y^*$ are primal and dual optimal solutions, then $c^Tx^* = b^Ty^*$.
\medskip

\parenproclaim Lemma F (Farkas lemma). For an arbitrary matrix $A$ and vector $b$, exactly one of the following holds:
\medskip
\item {i)} There exists a vector $x$ such that $Ax = b$ and $x \geq 0$.
\smallskip
\item {ii)} There exists a vector $y$ such that $A^Ty \geq 0$ and $b^Ty< 0$.
\medskip

\parenproclaim Lemma G (Farkas lemma 2.0). For an arbitrary matrix $A$ and vector $b$, exactly one of the following holds:
\medskip
\item {i)} There exists a vector $x$ such that $Ax \leq b$ and $x \geq 0$.
\smallskip
\item {ii)} There exists a vector $y$ such that $A^Ty \geq 0$, $y\geq0$ and $b^Ty< 0$.
\medskip

@@

Given a graph $G=(V,E)$, a {\it matching} is a set $M\subset E$ such that for all $v\in V$ there is at most one edge $e\in M$ with $v\in e$. The linear program that finds the maximal matching in a bipartite graph is
$$\matrix{
    & \maximise   & \quad & \displaystyle\sum_{e\in E} x_e  \hfill    & \hcr
    & \subject &       & \displaystyle\sum_{v\in e}x_e \leq 1  \hfill &\hbox{for all}\ v\in V,\hcr
    &                   &       & x_e \geq 0      \hfill          &\hbox{for all}\ e\in E.\hfill
}$$
The dual of this linear program is the following:
$$\matrix{
    & \minimise   & \quad & \displaystyle\sum_{v\in V} y_v  \hfill & \hcr
    & \subject &       & y_u + y_v \geq 1   \hfill &\hbox{for all}\ \{u,v\}\in E,\hcr
    &                   &       & y_v \geq 0        \hfill  &\hbox{for all}\ v\in V.\hfill
}$$
    Given a graph $G=(V,E)$, a {\it vertex cover} is a set $S \subset V$ such that for all $\{u, v\}\in E$, $\{u,v\}\cap S\geq 1$. The dual linear program above finds the minimum vertex cover in a bipartite graph. The duality of these two programs results in K\H{o}nig's theorem.

\parenproclaim Theorem K (K\H{o}nig's theorem). Suppose $G=(V,E)$ is a bipartite graph. Let $M\subset E$ denote the maximum matching and let $S\subset V$ denote the minimum vertex cover. Then $|M|=|S|$.\slug

@@

\parenproclaim Theorem R (Rank characterisation of matroids). Let $E$ be a finite set and suppose that $r : {\cal P}(E)\rightarrow \R$ satisfies the three peoperties of the rank function. Then $M=(E,I)$, with
$$I = \{Y \subset E : |Y| = r(Y)\},$$
is a matroid with $r$ as its rank function.

@@

\parenproclaim Theorem E (Greedy equivalence). Let $M=(E,I)$ be an independence system. Then if the weighted greedy algorithm works for all (non-negative) weights, then $M$ is a matroid.\slug
